### 2018-06-04 doc/submission/reviewer-response.md

We'd like to thank the editors and the reviewers for their thorough and
insightful review of our manuscript. We have been able to make the vast majority
of the suggested revisions, which we feel has improved the manuscript considerably.
Below we present a point-by-point response to each of the editors' and reviewers' 
comments.

# Editor comment

- All three reviewers agreed that the paper had the strength of longitudinal
followup and would be of interest to readers of the JLR but they also noted
that there have been earlier studies with similar outcomes and one reviewer
noted that some of those papers were by the present group but were not
mentioned here. The reviewers also thought that the discussion was too
speculative and the authors need therefore, to moderate their enthusiasm. My
additional comment is that although you quote one paper by Yki-Jarvinen and
colleagues. you did not discuss/reference her more recent papers linking
shorter chain saturated fatty acids in the liver to NAFLD/NASH. You should look
at those papers and add them to the discussion. I look forward to a modified
version.

    *Thank you for this suggestion. We have made additions to the manuscript based on the editor's and reviewers'
    suggestions, including the discussion around SFA and NAFLD by
    Yki-Jarvinen and colleagues. We believe this has strengthened the paper and hope it
    satisfactorily addresses the editor and reviewers comments.*

### 2018-06-05 doc/submission/reviewer-response.md

<!--
The editorial staff will be in touch shortly to suggest revisions to your title
and abstract for you to consider. Please wait to hear from them before
submitting your revised manuscript. 
-->

# Editor comment

- All three reviewers agreed that the paper had the strength of longitudinal
followup and would be of interest to readers of the JLR but they also noted
that there have been earlier studies with similar outcomes and one reviewer
noted that some of those papers were by the present group but were not
mentioned here. The reviewers also thought that the discussion was too
speculative and the authors need therefore, to moderate their enthusiasm. My
additional comment is that although you quote one paper by Yki-Jarvinen and
colleagues. you did not discuss/reference her more recent papers linking
shorter chain saturated fatty acids in the liver to NAFLD/NASH. You should look
at those papers and add them to the discussion. I look forward to a modified
version.

    *Thank you for this suggestion. We have made additions to the manuscript based on the editor's and reviewers'
    suggestions, including the discussion around SFA and NAFLD by
    Yki-Jarvinen and colleagues (first paragraph, page 13). We believe this has strengthened the paper and hope it
    satisfactorily addresses the editor and reviewers comments.*

@@

2. The associations are powerful and stand alone. The authors have no data
directly implicating de novo lipogenesis, so the last sentence in the abstract
about previous research should be removed or substantially modified.

    *We thank the reviewer for these comments. We agree with this suggestion and
    have removed the sentence from the abstract.*

3. There are statements in the Discussion that are overly speculative, dated,
and misleading. At the bottom of page 13 the phrase "may lead to greater DNL in
an attempt to control blood glucose" should be removed. On page 14, lines 8-9
the sentence "Higher insulin resistance may encourage greater DNL to handle the
higher blood glucose" should be removed. In the next sentence on the same page,
the authors discuss reference 47 by implying that weight loss improved fatty
acid profiles through effects on DNL, but that is facile and does not consider
the complex effects of improved insulin sensitivity on lipoprotein clearance
associated with induction of lipoprotein lipase, so this sentence needs to be
eliminated or modified.

    *These concerns and suggestions are appropriately raised. We have rephrased
    these sentences at all points in the Discussion to be more cautious and less speculative.*

@@

    *We thank the reviewer for noticing those omissions. The r values have been
    included in the text.*

# Reviewer 2

1.	It is a frustrating that this is seemingly the third in a series of papers
that is looking at patterns in lipid fractions and association with metabolic
outcomes. Thus, the TG, NEFA and phospholipid fractions all seem to be a 'least
publishable unit'. The authors must integrate the knowledge gained from the
previous analysis of NEFA and phospholipids and discuss some of the seemingly
contradictory findings from the previous studies. If the data were examined,
would the DNL story 'hold up'?

    *We thank the reviewer for these comments. We agree that ideally the results
    of all the fatty acid fractions would have been included together, but given the number of
    fatty acids, the longitudinal visits, the depth and volume of the analyses,
    and the complex biology involved, it simply wasn't possible to contain
    everything in a single journal paper given word count and table/figure limitations. We have included a brief discussion of
    the results from the other fractions in the Discussion section (end of page 14, beginning of page 15). The 
    complete and integrated discussion and presentation of the results is found
    in the first author's (LWJ) PhD thesis (http://hdl.handle.net/1807/80893). As to a combined analysis,
    we have presented these results at the European Diabetes Epidemiology Group
    Meeting in April, 2018 (https://doi.org/10.6084/m9.figshare.6159293), and the proposed DNL mechanism
    remained clearly associated and had a strong contribution to associations
    with diabetes pathogenesis compared to other fatty acids/compositions.
    A short aside, these currently unpublished findings do not invalidate or 
    undermine the previous work and the current manuscript. The combined analysis uses a powerful
    statistical technique (PLS), but it has a number of major limitations,
    not least of which is it can only handle cross-sectional data and can not
    adjust for potential confounders. The present analysis in this manuscript uses PLS in addition
    to the powerful longitudinal technique GEE.*

@@

2.	The discussion of GEE and its limitations is well done. However, it seems
that the analysis suggests that total triglycerides are predictive of HOMA2-%S
and the other parameters. Do we need to subfractionate to get insight into risk
for type 2 diabetes?

    *This is excellent question. For a couple of reasons, no we don't need to 
    subfractionate TAG to understand risk. Firstly, we found that the four TGFA
    also highly positively correlated with clinical TAG. Secondly, clinical TAG
    had a much higher magnitude of association compared to any individual
    TGFA. Measuring clinical TAG is enough to assess risk. However, our results
    provide insight into the potential mechanisms regarding how TAG influences risk
    for diabetes, which we show could be due to higher amounts of the four TGFA
    and that higher amounts of these four TGFA alone may strongly contribute to
    the higher clinical TAG.*

3.	Supplemental figures 2 and 3 are not discussed at all. They appear to be
interesting and informative in the development of the model. While the process
is described in the earlier NEFA paper, the authors should help the reader
understand what information is being provided without having to go back to a
previous paper.

    *We thank the reviewer for this comment. We agree, this information is useful
    and have added a brief discussion of it to the Methods section (end of page 7).*

@@

# Reviewer 3

1. The authors noted that cluster of four TGFA that correlated negatively with
IS (14:0, 16:0, 14:1n-7, 16:1n-7) are products of DNL, and are therefore likely
reflecting a connection between DNL, CHO intake and diabetes risk. This
conclusion seems a bit overstated. Clearly, a link between this cluster and DNL
is possible, but the signature could be also be reflecting dietary intake,
hepatic steatosis and/or TGFA clearance by tissues such as muscle and adipose
tissue. The authors should include a more balanced discussion of potential
explanations/mechanisms beyond DNL.

    *We thank the reviewer for these comments. We have revised the discussion throughout to 
    include alternative possibilities.*

@@

2. Are there any additional analyses that can be offered to strengthen evidence
the TGFA composition is indeed a biomarker of DNL and/or CHO intake is this
specific cohort? For example, did the PROMISE study include food recall
records?

    *PROMISE did collect food frequency questionnaires, however this was done at
    the 3rd year visit while the fatty acids where measured at the baseline visit.
    So while we could technically use these dietary intake data to help understand the fatty acid data,
    we would likely obtain inaccurate or biased results because of the time separation.
    We therefore decided to not analyze these data.*

3. Also, the discussion should mention and consider that blood samples were
acquired after an overnight fast. Does fasting/fed state impact the correlation
between TGFA composition and DNL?

    *Excellent question. In the fed state, DNL would increase due to conversion
    of excess carbohydrates to fatty acids. This in turn would impact the composition
    of primarily the TGFA. In order to control for this source of variability
    in our cohort we only measured TGFA during fasting. We have included a brief
    description of this fed vs fasted state in the limitations section (middle of page 15).*

@@

Minor comments.

4. The authors should comment on how results reported here combine with
(consistencies/contradictions) those reported in their recent Diabetalogia paper
(the parallel NEFA analysis).

    *We have included a comment in the discussion on how our current analysis
    fits with our previous NEFA analysis (end of page 14, beginning of page 15).*

@@

5. Authors underscore the strength of the statistical modeling used for the
analysis, including the use DAG modeling to identify potential confounders.
Readers might benefit from a brief comment on the basic principles and strengths
of the method.

    *We agree with this suggestion, but initially excluded it from the methods to
    limit word count. We have included a section in the methods that briefly 
    explains this method (last paragraph of page 7).*

### 2018-06-05 doc/manuscript.Rmd
minor edits and typo fixes in manuscript after final proofreading

TGFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be 
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of triheptadecanoin (17:0; Nu-Chek Prep, Inc Elysian, MN, USA) was added as an internal
standard prior to extracting total lipids according to the method of Folch et al
[@Folch1957a]. Each serum lipid fraction (non-esterified fatty acids (NEFA), 
cholesteryl ester, phospholipid, and TG) was isolated using thin layer 
chromatography. Fatty acid methyl esters were separated and quantified using a 
Varian-430 gas chromatograph (Varian, Lake Forest, CA, USA) equipped with a 
Varian Factor Four capillary column and a flame ionization detector. Fatty acid 
concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a]. There
were 22 fatty acids measured in the TGFA fraction. 
Findings for other lipid fractions in this cohort are reported separately (see
ref [@Johnston2016a] for the phospholipid and cholesteryl ester fraction and
[@Johnston2017b] for the NEFA fraction analysis).

@@

Generalized estimating equation (GEE) models [@Zeger1986a] were used in the 
primary analysis to determine the longitudinal associations between the outcome 
variables and the predictor variables. The predictor variables
were scaled (mean-centered and standardized). Given the longitudinal 
design, an auto-regressive of order 1 working correlation matrix was 
specified in the GEE model. Covariates to adjust for were selected based on the
previous literature, from directed acyclic graph (DAG) [@Greenland1999a]
recommendations, and from quasi-likelihood information criteria. 
DAGs are used to identify the minimum adjustment necessary for a model by
using the causal pathways to algorithmically identify potential confounding and
colliding variables (see [@Greenland1999a] for more detail about using
DAGs). The DAG structures to understand potential confounding, shown in 
`r cite_sf("dag_is")` and `r cite_sf("dag_bcf")`, were processed by the DAGitty
software [@Textor2011a;@Shrier2008a] to generate the recommended adjustments.
These DAG structures were developed based on hypothesized causal pathways
between each variable, which were then input into the DAGitty software. The
output from DAGitty was used to help inform the final model.

@@

There are a few possible explanations for these findings. Circulating 
TGFA derive from three sources: adipose lyposis, dietary fat, and de novo
lipogenesis (DNL). Dietary carbohydrates and fat can influence DNL activity
[@Aarsland1998a;@Harding2015a;@Luukkonen2018a]. Determining the specific source
of TGFA is extremely difficult to ascertain outside of highly controlled
experimental settings. Many previous studies that have examined DNL
have used markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or
16:1n-7 to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are
limitations to using these ratios as the fatty acids used in their calculation
can also be obtained from the diet in addition to being created through DNL
[@Hodson2008a]. An experimental feeding trial (n=24) was conducted to identify the fatty acids
that most accurately reflected DNL as potential biomarkers [@Lee2015a]. The
study found that palmitoleic acid (16:1n-7), directly measured DNL using
isotopes, and liver fat were all highly correlated with each other (r>0.50),
suggesting that 16:1n-7 may be a good biomarker for hepatic DNL. In another
small (n=14) feeding trial, meal type (high fat vs low fat) was tested to
determine its effect on DNL and TGFA composition [@Wilke2009a]. The authors
reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
carbohydrate) group. In another recent overfeeding trial, 1000 kcal of saturated
fat, unsaturated fat, or carbohydrates over 3 weeks was given to 38 overweight
individuals [@Luukkonen2018a] to test changes in liver fat and hepatic DNL. At the
end of the study, the carbohydrate group had higher DNL as well as an increase
in liver fat, though the saturated fat group had the highest increase in liver
fat.

This link between carbohydrate intake and DNL activity has been well documented
[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a; @Chehade2013a; @Verges2015a].
In our findings, the four fatty acids were highly positively correlated
amongst each other and negatively or neutrally with all other TGFA, in addition to
clustering together on their negative association with insulin sensitivity. This
may suggest that greater DNL activity may be the source of these TGFA. Several
studies have shown a link between higher estimated DNL and an increased risk for
metabolic dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]; how
DNL may influence metabolic dysfunction is not well understood. Possible
reasons may be that higher DNL produces more of certain fatty acids or that
higher DNL increases circulating TG, which itself is well documented to
contribute to metabolic dysfunction and which we found in our study of a high
positive correlation between the four TGFA and clinical TG.

Regardless of the exact source of these fatty acids, our results, in addition to
the available scientific evidence, emphasize the importance of the fatty acid
composition on metabolic health, as individual fatty acids can have specific
physiological functions. For instance, a higher concentration of
circulating 14 and 16 carbon fatty acids may expose tissues to greater
lipotoxicity, for instance from palmitic acid (16:0), which is well-known to
have harmful effects on tissues [@Riserus2008a; @Iggman2010a]. Our study extends
these findings by showing that TGFA with 14 to 16 carbons clustered together and
this pattern strongly associated with lower insulin sensitivity. While these fatty
acids also had a significant association with beta-cell function, the magnitude
of associations were more modest compared to those for insulin sensitivity.

The direction of association between TGFA and insulin sensitivity is unclear
from previous cross-sectional studies due to the physiological feedback
mechanisms involved. For example, while higher TGFA may promote
muscle insulin resistance, the reverse may also be true [@Flannery2012a].
As we found no interaction by time of TGFA on insulin sensitivity, this study
cannot determine the exact role of the feedback mechanism. However, combining the
with the lack of time interaction and the consistent negative association
without the time interaction, these results at least suggest that the feedback
mechanism may not be strongly influential and that TGFA may predict insulin
sensitivity at least over a six year period. Given the complex biological mechanisms and
feedback loops involved, disentangling whether insulin sensitivity influences
TGFA more strongly than TGFA influencing insulin sensitivity will require more
complex research designs and analyses.

Given the close biological relationship between circulating NEFA and TG, NEFA 
may act as confounding factor and was thus adjusted for. In our published
analysis of the NEFA fraction [@Johnston2017b], we found that higher total NEFA,
but not the specific composition, associated with lower beta-cell function. This
is in contrast to the TGFA findings that the specific composition does
differentially associate with insulin sensitivity and beta-cell function,
adjusting for total NEFA. There was no difference in results in models that did
not include NEFA as a confounder (data not shown). This difference in results
between NEFA and TGFA suggests that TGFA may independently and strongly influence the 
pathophysiology of T2D, when compared to other lipid fraction compositions,
including the phospholipid and cholesteryl ester fractions [@Johnston2016a]. This
may be because TG are biologically destined for uptake by non-hepatic tissue as
they are found mainly in VLDL, at least during fasting. This is in contrast to
NEFA that are mostly taken up by the liver and used in TG production [@Nielsen2012a].

@@

>>>>

TGFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be 
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of triheptadecanoin (17:0; Nu-Chek Prep, Inc Elysian, MN, USA) was added as an internal
standard prior to extracting total lipids according to the method of Folch et al
[@Folch1957a]. Each serum lipid fraction (non-esterified fatty acids (NEFA), 
cholesteryl ester, phospholipid, and TG) was isolated using thin layer 
chromatography. Fatty acid methyl esters were separated and quantified using a 
Varian-430 gas chromatograph (Varian, Lake Forest, CA, USA) equipped with a 
Varian Factor Four capillary column and a flame ionization detector. Fatty acid 
concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a]. There
were 22 fatty acids measured in the TGFA fraction. 
Findings for other lipid fractions in this cohort are reported separately (see
[@Johnston2016a] for the phospholipid and cholesteryl ester fraction and
[@Johnston2017b] for the NEFA fraction analysis).

@@

Generalized estimating equation (GEE) models [@Zeger1986a] were used in the 
primary analysis to determine the longitudinal associations between the outcome 
variables and the predictor variables. The predictor variables
were scaled (mean-centered and standardized). Given the longitudinal 
design, an auto-regressive of order 1 working correlation matrix was 
specified in the GEE model. Covariates to adjust for were selected based on the
previous literature, from directed acyclic graph (DAG) [@Greenland1999a]
recommendations, and from quasi-likelihood information criteria. 
DAGs are used to identify the minimum adjustment necessary for a model by
using the causal pathways to algorithmically identify potential confounding and
colliding variables (see [@Greenland1999a] for more detail about using
DAGs). The DAG structures to understand potential confounding, shown in 
`r cite_sf("dag_is")` and `r cite_sf("dag_bcf")`, were processed by the DAGitty
software [@Textor2011a;@Shrier2008a] to generate the recommended adjustments.
These DAG structures were developed based on hypothesized causal pathways
between each variable, which were then input into the DAGitty software. The
output from DAGitty was used, in conjunction with the other methods, to help inform the final model.

@@

There are a few possible explanations for these findings. Circulating 
TGFA derive from three sources: adipose lipolysis, dietary fat, and de novo
lipogenesis (DNL). Dietary carbohydrates and fat can influence DNL activity
[@Aarsland1998a;@Harding2015a;@Luukkonen2018a]. Determining the specific source
of TGFA is extremely difficult to ascertain outside of highly controlled
experimental settings. Many previous studies that have examined DNL as the source
have used markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or
16:1n-7 to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are
limitations to using these ratios as the fatty acids used in their calculation
can also be obtained from the diet in addition to being created through DNL
[@Hodson2008a]. An experimental feeding trial (n=24) was conducted to identify the fatty acids
that most accurately reflected DNL as potential biomarkers [@Lee2015a]. The
study found that palmitoleic acid (16:1n-7), directly measured DNL using
isotopes, and liver fat were all highly correlated with each other (r>0.50),
suggesting that 16:1n-7 may be a good biomarker for hepatic DNL. In another
small (n=14) feeding trial, meal type (high fat vs low fat) was tested to
determine its effect on DNL and TGFA composition [@Wilke2009a]. The authors
reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
carbohydrate) group. In another recent overfeeding trial, 1000 kcal of saturated
fat, unsaturated fat, or carbohydrates over 3 weeks was given to 38 overweight
individuals [@Luukkonen2018a] to test changes in liver fat and hepatic DNL. At the
end of the study, the carbohydrate group had higher DNL activity as well as an increase
in liver fat, though the saturated fat group had the highest increase in liver
fat. This link between carbohydrate intake and DNL activity has been well documented
[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a; @Chehade2013a; @Verges2015a].

In our findings, the four fatty acids were highly positively correlated
amongst each other and negatively or neutrally with all other TGFA, in addition to
clustering together on their negative association with insulin sensitivity. This
may suggest that greater DNL activity is the source of these TGFA. Several
studies have shown a link between higher estimated DNL activity and an increased risk for
metabolic dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]. How
DNL may influence metabolic dysfunction is not well understood. Possible
reasons may be that higher DNL produces more of certain fatty acids or that
higher DNL increases circulating TG, which itself is well documented to
contribute to metabolic dysfunction and which we found in our study as a high
positive correlation between the four TGFA and clinical TG.

Regardless of the exact source of these fatty acids, our results, in addition to
the available scientific evidence, emphasize the importance of the fatty acid
composition on metabolic health, as individual fatty acids can have specific
physiological functions. For instance, a higher concentration of
circulating 14 and 16 carbon fatty acids may expose tissues to greater
lipotoxicity, for instance from palmitic acid (16:0), which is well-known to
have harmful effects on tissues [@Riserus2008a; @Iggman2010a]. Our study extends
these findings by showing that TGFA with 14 to 16 carbons clustered together and
this pattern strongly associated with lower insulin sensitivity. While some of these fatty
acids also had a significant association with beta-cell function, the magnitude
of associations were more modest compared to those for insulin sensitivity.

The direction of association between TGFA and insulin sensitivity is unclear
from previous cross-sectional studies due to the physiological feedback
mechanisms involved. For example, while higher TGFA may promote
muscle insulin resistance, the reverse may also be true [@Flannery2012a].
As we found no interaction by time of TGFA on insulin sensitivity, this study
cannot determine the exact role of the feedback mechanism. However, combining
the lack of a time interaction and the consistent negative association
in models without the time interaction, these results at least suggest that the feedback
mechanism may not be strongly influential and that TGFA may predict insulin
sensitivity at least over a six year period. Given the complex biological mechanisms and
feedback loops involved, disentangling whether insulin sensitivity influences
TGFA more strongly than TGFA influencing insulin sensitivity will require more
complex research designs and analyses.

Given the close biological relationship between circulating NEFA and TG, NEFA 
may act as a confounding factor and was thus adjusted for. In our published
analysis of the NEFA fraction [@Johnston2017b], we found that higher total NEFA,
but not the specific composition, associated with lower beta-cell function. This
is in contrast to the TGFA findings that the specific composition does
differentially associate with insulin sensitivity and beta-cell function,
adjusting for total NEFA. There was no difference in results in models that did
not include NEFA as a confounder (data not shown). This difference in results
between NEFA and TGFA suggests that TGFA may independently and strongly influence the 
pathophysiology of T2D, when compared to other lipid fraction compositions,
including the phospholipid and cholesteryl ester fractions [@Johnston2016a]. This
may be due to TG being biologically destined for uptake by non-hepatic tissue as
they are found mainly in VLDL, at least during fasting. This is in contrast to
NEFA that are mostly taken up by the liver and used in TG production [@Nielsen2012a].

@@

Note: Tables values as either median (interquartile range), mean (SD), or n (percent).
